Exploring the tree of numerical semigroups

### 2014-11-03 Papier/article.tex
Hal version

\title{Exploring the tree of numerical semigroups}
\author{Jean Fromentin and  Florent Hivert}
\date{2 October 2014}
\keywords{semigroups, tree, algorithm, \texttt{SSE}, 
\texttt{Cilk}, optimization}
\subjclass[2010]{05A15,68R05,68W10}

@@

\begin{abstract}
  In this paper we describe an algorithm visiting all the numerical semigroups
  up to a given genus using a new representation of numerical semigroups. The
  interest of this algorithm is that it fits particularly well the
  architecture of modern computer allowing very large optimizations : we obtain the number of numerical semigroups of genus $g\leq  67$ and we confirm the Wilf conjecture for $g\leq 60$.

@@

A \emph{numerical semigroup} $S$ is a subset of $\NN$ containing $0$, close under addition and of finite complement in $\NN$.  
For example the set 
\begin{equation}
\label{E:NSG}
S_E=\{0,3,6,7,9,10\}\cup\{x\in\NN, x\geq 12\}
\end{equation}
is a numerical semigroup.
The \emph{genus} of a numerical semigroup $S$, denoted by~$g(S)$, is the cardinality of~$\NN\setminus S$.
 For example the genus of $S_E$ is $6$,  the cardinality of $\{1,2,4,5,8,11\}$

For a given positive integer $g$, the number of numerical semigroups of genus $g$ is finite and is denoted by $n_g$. 
In  J.A. Sloane's \emph{on-line encyclopedia of integer}~\cite{OEIS} we find the values of $n_g$ for $g\leq 52$. 
These values have been obtain by M. Bras-Amor\'os (view \cite{BrasAmoros2008} for more details for $g\leq 50)$. 
On his home page~\cite{Delgado}, M. Delgado  gives the value of $n_{55}$ without specifying the values of $n_{53}$ and~$n_{54}$.

M.~Bras-Amor\'os used a depth first search exploration of the tree of numerical semigroups $\mathcal{T}$ up to a given genus.
This tree was introduced by J.C.~Rosales and al. in \cite{Rosales} and it is the subject of the Section~\ref{S:Tree}.
Starting with all the numerical semigroups of genus $49$ she obtained the number of numerical semigroups of genus $50$ in $18$ days on a pentium D runing at $3$GHz. 
In the package~\tt{NumericalSgs} \cite{NumericalSgps} of \tt{GAP} \cite{GAP}, M. Delgado together with P.A.~Garcia-Sanchez and J. Morais used the same method of exploration.

Here we describe a new algorithm for the exploration of the tree of numerical
semigroup $\mathcal{T}$ and achieve the computation of $n_g$ for $g\leq 67$.
The cornerstone of our method is a new combinatorial representation of
numerical semigroup that is well suited for exploration of the
tree~$\mathcal{T}$ and allow large code optimization essentially based on use of
vectorial instructions and parallelization.
The goal of the paper is twofold:
first to present our new encoding of the semigroup and the associated
algorithms, and second to present the optimization techniques which allows, for
those kinds of algorithms, to get speedups by factors of hundreds and even
thousands. We claim that these technique are fairly general for those kinds of
algorithms. As a support for the claim, we applied it to an algorithm of
N.~Borie enumerating integer vector modulo permutation groups~\cite{Borie} and
got a speedup by a factor larger than 2000 using 8 cores.

The paper is divided as follows.
In Section~\ref{S:Tree} we describe the tree of numerical semigroups and give bounds for some parameters attached to a numerical semigroup.
The decription of our new representation of numerical semigroup is done in the third section.
In Section~\ref{S:Algo} we describe an algorithm based on the representation given in Section~\ref{S:DecNumber} and give its complexity. 
Section~\ref{S:Opti} is more technical, and is devoted to the optimisation of the algorithm introduced in Section~\ref{S:Algo}.
In the last section we emphase the results obtained using our algorithm.

@@

We start this section with definitions and properties of numerical semigroups that will be used in the sequel.
For a more complete introduction, the reader can usefully consults the book \emph{Numerical Semigroups} of J.C.~Rosales and P.A.~Garc\'ia-S\'anchez \cite{BookNS} or the book \emph{The Diophantine Frobenius Problem} of J.L.~Ram\'irez~Alfons\'in \cite{BookDFP}.

@@

By definition a numerical semigroup is an infinite object and we need a finite description of such a object. 
That is the role of generating sets.

@@

If we reconsider the numerical semigroup of \eqref{E:NSG}, we obtain
\begin{equation}
\label{E:GNSG}
S_E=\{0,3,6,7,9,10\}\cup[12,+\infty[=\left<3,7\right>
\end{equation}

A non-zero element $x$ of a numerical semigroup $S$ is said to be \emph{irreducible} if it cannot be expressed as a sum of two non-zeros elements of $S$.  
We denote by $\Irr(S)$ the set of all irreducible elements of $S$.

@@

\begin{proof}
$i)$ Let $x$ ba a positive integer satisfying $x \geq c(S)+m(S)$. 
As the integers $x$ and $x-m(S)$ are greater than $c(S)$, they belong to $S$.
Since $x$ equals $(x-m(S))+m(S)$ with $x-m(S)\not=0$ and $m(S)\not=0$, it could not be irreducible.

@@

$iii)$ Let $x$ be an element of $S\cap\{0,...,c(S)-1\}$. 
The integer $y=c(S)-1-x$ lies in $\{0,...,c(S)-1\}$. 
Moreover $y$ is not in $S$, for otherwise  we write $c(S)-1=y+x$ where $x,y$ are elements of $S$ implying $c(S)-1\in S$. 
In contradiction with the definition of conductor. 
Thus we define an involution 
\[
\begin{array}{rcl}
\psi:\{0,...,c(S)-1\}&\to&\{0,...,c(S)-1\}\\
x&\mapsto&c(S)-1-x
\end{array}
\]
that send $S\{0,...,c(S)-1\}$ into $\{0,...,c(S)-1\}\setminus S$ which is  of cardinality~$g(S)$. 
Let us denote by $k$ the cardinality  of $S\cap\{0,...,c(S)-1\}$. 
Since $\psi$ is injective we must have $k\leq g(S)$. 
From the relation 
\[
\{0,...,c(S)-1\}=S\cap\{0,...,c(S)-1\}\sqcup\{0,...,c(S)-1\}\setminus S
\]
 we obtain $c(S)=k+g(S)$ and so $c(S)\leq 2 g(S)$.
\end{proof}

A consequence of Proposition~ \ref{P:Res}~$i)$ is that $\Irr(S)$ is finite and its cardinality is at most $c(S)+m(S)-1$.
More precisely, using Ap\'ery set, we can prove that  the cardinality of $\Irr(S)$ is at most $m(S)$. See Section~2 of Chapter I of~\cite{BookNS} for more details.

@@

Proposition~\ref{P:Sx} implies that every semigroup $S$ of genus $g$ can be obtained from a semigroup $S$ by removing a generator $x$ of $S$ that is greater than $c(S)$.
Hence the relation~$S'=S^x$ holds.

We construct the tree of numerical semigroups, denoted by $\mathcal{T}$ as follows. 
The root of the tree is the unique semigroup of genus $0$, \ie, $\left<1\right>$ that is equal to $\NN$. 
If~$S$ is a semigroup in the tree,  the sons of $S$ are exactly the semigroup~$S^x$ where $x$ belongs to $\Irr(S)\cap[c(s),+\infty[$. 
By convention, when depicting the tree, the numerical semigroup $S^x$ is in the left of $S^y$ if $x$ is smaller than~$y$. 
With this construction, a semigroup $S$ has depth $g$ in $\mathcal{T}$ if and only if its genus is~$g$, see Figure~\ref{F:Tree}.
We denote by $\mathcal{T}_{g}$ the subtree of~$\mathcal{T}$ resticted to all semigroup of genus $\leq g$.

 \begin{figure}[hf!]
\[
\begin{tikzpicture}[level distance=2cm, inner sep=2mm,level/.style={sibling distance=1.55cm}]
    \node {$\sgnode{1}$}
    child {node {$\sgnode{2,3}$}
      child [sibling distance=4.5cm] {node {$\sgnode{3,4,5}$}
        child {node {$\sgnode{4,5,6,7}$}
          child [sibling distance=2cm]{node {$\sgnode{5,6,7,8,9}$}  edge from parent node [left] {4}}
          child [sibling distance=1.9cm]{node {$\sgnode{\gr 4,6,7,9}$}  edge from parent node [left] {5}}
          child {node {$\sgnode{\gr 4,\gr 5,7}$}  edge from parent node [right] {6}}
          child {node {$\sgnode{\gr 4,\gr 5,\gr 6}$}  edge from parent node [right] {7}}
          edge from parent node [left] {3}
        }
        child{{} edge from parent[draw=none]}
        child{{} edge from parent[draw=none]}
        child {node {$\sgnode{\gr 3,5,7}$}
          child {node {$\sgnode{\gr 3,7,8}$} edge from parent node [left] {5}}
          child {node {$\sgnode{\gr 3,\gr 5}$} edge from parent node [right] {7}}
          edge from parent node [left] {4}
        }
        child {node {$\sgnode{\gr 3,\gr 4}$}
          edge from parent node [right] {5}
        }
        edge from parent node [left] {2}
      }
      child [sibling distance=4.5cm] {node {$\sgnode{\gr2,5}$}
        child {node {$\sgnode{\gr2,7}$}
          child {node {$\sgnode{\gr2,9}$}
            edge from parent node [right] {7}
          }
          edge from parent node [right] {5}
        }
        edge from parent node [right] {3}
      }
      edge from parent node [left] {1}
    }
    ;
  \end{tikzpicture}
  \]
\caption{The first four layers of the tree $\mathcal{T}$ of numerical semigroups, corresponding to~$\mathcal{T}_4$. A generator of a semigroup is in gray if is not greater than $c(S)$. An edge between a semigroup $S$ and its son $S'$ is labelled by  $x$ is $S'$ is obtained from $S$ by removing~$x$, that is $S'=S^x$.}
\label{F:Tree}
\end{figure}

As the reader can check, the main difficulty to characterize the son of a semigroup is to determine its irreducible elements.
 In \cite{NumericalSgps}, the semigroup are given by their Ap\'ery set and then the difficulty is to describe the Ap\'ery set of $S^x$  from this of $S$. 
 This approach is elegant but not sufficiently \emph{basic} for our optimisations.

@@

Assume that $y$ is an element of $D_S(x)$.
By very definition of $D_S(x)$, the integer~$z=x-y$ also belongs to $S$. 
Then $x$ can be decomposed as $x=y+z$ with $y$ and $z$ in~$S$. 
Moreover the condition $2y\leq x$ implies $y\leq z$. 
In other words if we define~$D'_S(x)$ to be the set of all $(y,z)\in S\times S$ with $x=y+z$ and $y\leq z$ then~$D_S(x)$ is the image of $D'_S(x)$ under the projection on the first coordinate.
 Hence~$D_S(x)$ describes how~$x$ can be decomposed as sums of two elements of $S$.
 This is why we decide to call $d_S(x)$ the $S$-decomposition number of $x$.

\begin{exam}
Reconsider the semigroup $S_E$ given at \eqref{E:NSG}.
The integer $14$ admits three decompositions as sums of two elements of $S$, namely $14=0+14$ and $14=7+7$. 
Thus the set $D_{S_E}(14)$ is equal to $\{0,7\}$ and $d_{S_E}$ equals $2$.
\end{exam}

@@

Let us show $ii)$. 
Assume $x$ is irreducible in $S$.
There cannot exist $y$ and $z$ in~$(S\setminus\{0\})^2$ such that~$x=y+z$. 
The only possible decomposition of $x$ as a sum of two elements of $S$ is $x=0+x$. 
Hence, the set $D_S(x)$ is equal to~$\{0\}$ and we have $d_S(x)=1$. 
Conversely, let $x$ such that $d_S(x)$ is $1$. 
By~$i)$ the  integer $x$ must be in $S$. 
As $x=0+x$ is always a decomposition of $x$ as a sum of two elements in $S$, we obtain $D_S(x)=\{0\}$.
If there exist $y$ and $z$ in $S$ such that $y\leq z$ and $x=y+z$ hold then $y$ lies in~$D_S(x)$. 
This implies $y=0$ and $z=x$. 
Hence $x$ is irreducible in~$S$.
\end{proof}

We note  that $0$ is never irreducible despite the fact $d_S(0)$ is $1$ for all numerical semigroup $S$.

We now explain how to compute the $S$-decomposition numbers function of a numerical semigroup  from this of its father.

@@

\begin{proof}
Let $y$ be in $\NN\setminus\{0\}$. We have 
\[
D_{S^x}(y)=\{z \in S^x\ |\ y-z\in S^x\ \text{and}\  2z\leq y\}.
\]
We construct two sets as follow :
\begin{align*}
E&=\{z\in S^x\ |\ y-z=x\ \text{and}\ 2z\leq y\}\\
F&=\begin{cases}
\{x\}&\text{if $y-x\in S$ and $2x\leq y$},\\
\emptyset&\text{otherwise}.
\end{cases}
\end{align*}
We have the relation $D_{S}(y)=D_{S^x}(y)\sqcup E\sqcup F$ and so $d_S(y)$ is equal to $d_{S^x}(y)+\card(E)+\card(F)$
and $D_{S^x}(y)$ is a subset of $D_S(y)$. We start by determining $E$. 
The relation $y-z=x$ implies $z=y-x$ and $2z\leq y$ becomes $y\leq 2x$. Thus we obtain
\[
E=\begin{cases}
\{y-x\}&\text{if $y-x\in S^x$ and $y\leq 2x$}\\
\emptyset&\text{otherwise}
\end{cases}
\] 
The relation $y-x\in S^x$ together with $y\leq2x$ is equivalent to $y-x\in S$ together with $y<2x$.
Therefore we obtain
\[
E=\begin{cases}
\{y-x\}&\text{if $y-x\in S$ and $y< 2x$}\\
\emptyset&\text{otherwise}
\end{cases}
\] 
Assume $y-x$ lies in $S$.  If $y<2x$ holds, we have $E=\{y-x\}$ and $F=\emptyset$. Otherwise, we obtain $E=\emptyset$ and $F=\{x\}$. 
Then if $y-x\in S$ holds, we have $\card(E)+\card(F)=1$ and so $d_{S}(y)$ is equal to $d_{S^x}(y)+1$.
If $y-x$ not belong to $S$, then $E$ and $F$ are empty implying $d_{S}(y)=d_{S^x}(y)$. As the condition $y-x\in S$ is equivalent to $y\geq x$ together with $d_S(y-x)>0$, by Proposition~\ref{P:Dandd} the result holds.
\end{proof}

@@

In Algorithm~\ref{A:ExpRec} we do not specify how to compute $c(S)$, $g(S)$ and $m(S)$  from $S$ neither how to test if an integer is irreducible. 
It also miss  the characterisation of $S^x$ from $S$.
These items depend heavily of the representation of $S$.
Our choice is to use the $S$-decomposition numbers function. 
The first task is to use a finite set of such numbers to characterise the whole semigroup.

\begin{prop}
\label{P:delta}
Let $G$ be an integer and $S$ be a numerical semigroup of genus $g\leq G$.
Then $S$ is entirely described by $\delta_S=(d_S(0),...,d_S(3\,G))$. More precisely we can obtain $c(S), g(S), m(S)$ and $\Irr(S)$ from $\delta_S$.
\end{prop}

\begin{proof}
By Proposition~\ref{P:Res} $iii)$ we have  the relation $c(S)\leq 2g(S)$ and so the $S$-decomposition number of~$c(S)$ occurs in $\delta_S$.
Since $c(S)$ is equal to $1+\max(\NN\setminus S)$ for $S\not=\NN$, Proposition~\ref{P:delta} implies 
\[
c(S)=1+\max\{i\in\{0,...,3G\},\ d_S(i)=0\}.
\]
As all elements of $\NN\setminus S$ are smaller than $c(S)$, their $S$-decomposition numbers are in $\delta_S$ and we obtain
\[
g(S)=\card\{i\in\{0,...,3G\},\ d_S(i)=0\}.
\]
By Proposition~\ref{P:Res} $ii)$, the relation $m(S)\leq g(S)+1$ holds. 
This implies that the $S$-decomposition number of $m(S)$ appears in $\delta_S$ : 
\[
m(S)=\min\{i\in\{0,...,3G\},\ d_S(i)>0\}.
\]
Since, by Proposition~\ref{P:Res} $i)$,  all irreducible elements are smaller than $c(S)+m(S)-1$, which is itself smaller than $3G$ from the relations $ii)$ and $iii)$ of Proposition~\ref{P:Res}. So Proposition~\ref{P:Dandd} gives
\[
\Irr(S)=\{i \in\{0,...,3G\},\ d_S(i)=1\}.
\]
\end{proof}

Even though it is quite simple, the computation of $c(S),m(S)$ and $g(S)$ from $\delta_S$ has a non negligible cost.
We represent a numerical semigroup $S$ of genus $g\leq G$ by $(c(S),g(S),c(S),\delta_S)$. 
In an algorithmic context, if the variable \texttt{S} stands for a numerical semigroup  we use:

@@

Though there are asymptotically faster algorithms than the one presented here,
thank to careful optimizations, we were able to compute $n_g$ for much larger
genus than before. This is due to the fact that our algorithm is particularly
well suited for the current processor architecture. In particular, it allows
to use parallelism at various scale (parallel branch exploration,
vectorization)...

To get the greatest speed from modern processor, we used several optimization
tricks, which we will elaborate in the following section:

@@

\subsection{Vectorization}

Assume for example that we want to construct the tree~$\mathcal{T}_{100}$
of all numerical semigroup of genus smaller than $100$.  In this case, the
representation of numerical semigroup given in Section~\ref{S:DecNumber} uses
decomposition numbers of integers smaller than~$300$.  By Lemma~\ref{L:RepN},
such a decomposition number is smaller than $151$ and requires~$1$ byte of
memory.  Thus at each \textbf{for} step of Algorithm~\textsc{Son}, the CPU
actually works on $1$ byte. However current CPUs usually work on $8$ bytes and
even on $16$~bytes using vector extensions. 
The first optimization uses this point.

>>>>

\title{Exploring the tree of numerical semigroups}
\author{Jean Fromentin and  Florent Hivert}
\date{3 November 2014}
\keywords{semigroups, tree, algorithm, \texttt{SSE}, 
\texttt{Cilk}, optimization}
\subjclass[2010]{05A15,68R05,68W10}

@@

\begin{abstract}
  In this paper we describe an algorithm visiting all numerical semigroups
  up to a given genus using a new representation. The
  interest of this algorithm is that it fits particularly well the
  architecture of modern computers allowing very large optimizations: we obtain the number of numerical semigroups of genus $g\leq  67$ and we confirm the Wilf conjecture for $g\leq 60$.

@@

A \emph{numerical semigroup} $S$ is a subset of $\NN$ containing $0$, closed under addition and of finite complement in $\NN$.  
For example the set 
\begin{equation}
\label{E:NSG}
S_E=\{0,3,6,7,9,10\}\cup\{x\in\NN, x\geq 12\}
\end{equation}
is a numerical semigroup.
The \emph{genus} of a numerical semigroup $S$, denoted by~$g(S)$, is the cardinality of~$\NN\setminus S$.
 For example the genus of $S_E$ is $6$,  the cardinality of $\{1,2,4,5,8,11\}$.

For a given positive integer $g$, the number of numerical semigroups of genus $g$ is finite and is denoted by $n_g$. 
In  J.A. Sloane's \emph{on-line encyclopedia of integer sequences}~\cite{OEIS} we find the values of $n_g$ for $g\leq 52$. 
These values have been obtained by M. Bras-Amor\'os (\cite{BrasAmoros2008} for more details for $g\leq 50)$. 
On his home page~\cite{Delgado}, M. Delgado  gives the value of $n_{55}$. 
The values of $n_{53}$ and $n_{54}$ seem to be so far available nowhere.

M.~Bras-Amor\'os used a depth first search exploration of the tree of numerical semigroups $\mathcal{T}$ up to a given genus.
This tree was introduced by J.C.~Rosales and al. in \cite{Rosales} and it is the subject of Section~\ref{S:Tree}.
Starting with all the numerical semigroups of genus $49$ she obtained the number of numerical semigroups of genus $50$ in $18$ days on a pentium D runing at $3$GHz. 
In the package~\tt{NumericalSgs} \cite{NumericalSgps} of \tt{GAP} \cite{GAP}, M.~Delgado together with P.A.~Garcia-Sanchez and J. Morais used the same method of exploration.

Here we describe a new algorithm for the exploration of the tree of numerical
semigroups $\mathcal{T}$ and achieve the computation of $n_g$ for $g\leq 67$.
The cornerstone of our method is a new combinatorial representation of
numerical semigroups that is well suited and allows large code optimization essentially based on the use of
vectorial instructions and parallelization.
The goal of the paper is twofold:
first to present our new encoding of numerical semigroups and the associated
algorithms, and second to present the optimization techniques which allow, for
those kinds of algorithms, to get speedups by factors of hundreds and even
thousands. We claim that these techniques are fairly general for those kinds of
algorithms. As a support for the claim, we applied it to an algorithm of
N.~Borie enumerating integer vector modulo permutation groups~\cite{Borie} and
got a speedup by a factor larger than 2000 using 8 cores.

The paper is divided as follows.
In Section~\ref{S:Tree} we describe the tree of numerical semigroups and give bounds for some parameters attached to a numerical semigroup.
The description of our new representation of numerical semigroups is done in the second section.
In Section~\ref{S:Algo} we describe an algorithm based on the representation given in Section~\ref{S:DecNumber} and give its complexity. 
Section~\ref{S:Opti} is more technical, and is devoted to the optimization of the algorithm introduced in Section~\ref{S:Algo}.
In the last section we emphasize the results obtained using our algorithm.

@@

We start this section with definitions and properties of numerical semigroups that will be used in the sequel.
For a more complete introduction, the reader can usefully consult the book \emph{Numerical Semigroups} of J.C.~Rosales and P.A.~Garc\'ia-S\'anchez \cite{BookNS} or the book \emph{The Diophantine Frobenius Problem} of J.L.~Ram\'irez~Alfons\'in \cite{BookDFP}.

@@

By definition a numerical semigroup is an infinite object and we need a finite description of such an object. 
That is provided by generating sets.

@@

If we reconsider the numerical semigroup of \eqref{E:NSG}, we obtain
\begin{equation}
\label{E:GNSG}
S_E=\{0,3,6,7,9,10\}\cup[12,+\infty[=\left<3,7\right>.
\end{equation}

A non-zero element $x$ of a numerical semigroup $S$ is said to be \emph{irreducible} if it cannot be expressed as a sum of two non-zero elements of $S$.  
We denote by $\Irr(S)$ the set of all irreducible elements of $S$.

@@

\begin{proof}
$i)$ Let $x$ be a positive integer satisfying $x \geq c(S)+m(S)$. 
As the integers $x$ and $x-m(S)$ are greater than $c(S)$, they belong to $S$.
Since $x$ equals $(x-m(S))+m(S)$ with $x-m(S)\not=0$ and $m(S)\not=0$, it cannot be irreducible.

@@

$iii)$ Let $x$ be an element of $S\cap\{0,...,c(S)-1\}$. 
The integer $y=c(S)-1-x$ lies in $\{0,...,c(S)-1\}$. 
Moreover $y$ is not in $S$, for otherwise  we write $c(S)-1=y+x$ where $x,y$ are elements of $S$ implying $c(S)-1$ belongs to $S$. 
In contradiction with the definition of conductor. 
Thus we define an involution 
\[
\begin{array}{rcl}
\psi:\{0,...,c(S)-1\}&\to&\{0,...,c(S)-1\}\\
x&\mapsto&c(S)-1-x
\end{array}
\]
that sends $S\{0,...,c(S)-1\}$ into $\{0,...,c(S)-1\}\setminus S$ which is  of cardinality~$g(S)$. 
Let us denote by $k$ the cardinality  of $S\cap\{0,...,c(S)-1\}$. 
Since $\psi$ is injective we must have $k\leq g(S)$. 
From the relation 
\[
\{0,...,c(S)-1\}=S\cap\{0,...,c(S)-1\}\sqcup\{0,...,c(S)-1\}\setminus S
\]
 we obtain $c(S)=k+g(S)$ and so $c(S)\leq 2 g(S)$.
\end{proof}

A consequence of Proposition~ \ref{P:Res}~$i)$ is that $\Irr(S)$ is finite and its cardinality is at most $c(S)+m(S)-1$.
Moreover, the cardinality of $\Irr(S)$ is at most $m(S)$ since any two distinct elements of $\Irr(S)$ cannot be congruent modulo $m(S)$. See Section~2 of Chapter I of~\cite{BookNS} for more details.

@@

Proposition~\ref{P:Sx} implies that every semigroup $S$ of genus $g$ can be obtained from a semigroup $S$ by removing a generator $x$ of $S$ that is greater than or equal to $c(S)$.
Hence the relation~$S'=S^x$ holds.

We construct the tree of numerical semigroups, denoted by $\mathcal{T}$ as follows. 
The root of the tree is the unique semigroup of genus $0$, \ie, $\left<1\right>$ that is equal to $\NN$. 
If~$S$ is a semigroup in the tree,  the sons of $S$ are exactly the semigroups~$S^x$ where $x$ belongs to $\Irr(S)\cap[c(s),+\infty[$. 
By convention, when depicting the tree, the numerical semigroup $S^x$ is in the left of $S^y$ if $x$ is smaller than~$y$. 
With this construction, a semigroup $S$ has depth $g$ in $\mathcal{T}$ if and only if its genus is~$g$, see Figure~\ref{F:Tree}.
We denote by $\mathcal{T}_{g}$ the subtree of~$\mathcal{T}$ restricted to all semigroups of genus $\leq g$.

 \begin{figure}[hf!]
\[
\begin{tikzpicture}[level distance=2cm, inner sep=2mm,level/.style={sibling distance=1.55cm}]
    \node {$\sgnode{1}$}
    child {node {$\sgnode{2,3}$}
      child [sibling distance=4.5cm] {node {$\sgnode{3,4,5}$}
        child {node {$\sgnode{4,5,6,7}$}
          child [sibling distance=2cm]{node {$\sgnode{5,6,7,8,9}$}  edge from parent node [left] {4}}
          child [sibling distance=1.9cm]{node {$\sgnode{\gr 4,6,7,9}$}  edge from parent node [left] {5}}
          child {node {$\sgnode{\gr 4,\gr 5,7}$}  edge from parent node [right] {6}}
          child {node {$\sgnode{\gr 4,\gr 5,\gr 6}$}  edge from parent node [right] {7}}
          edge from parent node [left] {3}
        }
        child{{} edge from parent[draw=none]}
        child{{} edge from parent[draw=none]}
        child {node {$\sgnode{\gr 3,5,7}$}
          child {node {$\sgnode{\gr 3,7,8}$} edge from parent node [left] {5}}
          child {node {$\sgnode{\gr 3,\gr 5}$} edge from parent node [right] {7}}
          edge from parent node [left] {4}
        }
        child {node {$\sgnode{\gr 3,\gr 4}$}
          edge from parent node [right] {5}
        }
        edge from parent node [left] {2}
      }
      child [sibling distance=4.5cm] {node {$\sgnode{\gr2,5}$}
        child {node {$\sgnode{\gr2,7}$}
          child {node {$\sgnode{\gr2,9}$}
            edge from parent node [right] {7}
          }
          edge from parent node [right] {5}
        }
        edge from parent node [right] {3}
      }
      edge from parent node [left] {1}
    }
    ;
  \end{tikzpicture}
  \]
\caption{The first four layers of the tree $\mathcal{T}$ of numerical semigroups, corresponding to~$\mathcal{T}_4$. A generator of a semigroup is it in gray if is not greater than $c(S)$. An edge between a semigroup $S$ and its son $S'$ is labelled by  $x$ if $S'$ is obtained from $S$ by removing~$x$, that is if $S'=S^x$ holds.}
\label{F:Tree}
\end{figure}

As the reader can check, the main difficulty to characterize the sons of a semigroup is to determine its irreducible elements.
 In \cite{NumericalSgps}, the semigroups are given by their Ap\'ery sets and then the difficulty is to describe the Ap\'ery set of $S^x$  from that of $S$. 
 This approach is elegant but not sufficiently \emph{basic} for our optimizations.

@@

Assume that $y$ is an element of $D_S(x)$.
By very definition of $D_S(x)$, the integer~$z=x-y$ also belongs to $S$. 
Then $x$ can be decomposed as $x=y+z$ with $y$ and $z$ in~$S$. 
Moreover the condition $2y\leq x$ implies $y\leq z$. 
In other words if we define~$D'_S(x)$ to be the set of all $(y,z)\in S\times S$ with $x=y+z$ and $y\leq z$ then~$D_S(x)$ is the image of $D'_S(x)$ under the projection on the first coordinate.
 Hence~$D_S(x)$ describes how~$x$ can be decomposed as sums of two elements of $S$.
 This justifies the name given to the function $d_S$.

\begin{exam}
Reconsider the semigroup $S_E$ given at \eqref{E:NSG}.
The integer $14$ admits two decompositions as sums of two elements of $S$, namely $14=0+14$ and $14=7+7$. 
Thus the set $D_{S_E}(14)$ is equal to $\{0,7\}$ and $d_{S_E}=2$ holds.
\end{exam}

@@

Let us show $ii)$. 
Assume $x$ is irreducible in $S$.
There cannot exist $y$ and $z$ in~$(S\setminus\{0\})^2$ such that~$x=y+z$. 
The only possible decomposition of $x$ as a sum of two elements of $S$ is $x=0+x$. 
Hence, the set $D_S(x)$ is equal to~$\{0\}$ and we have $d_S(x)=1$. 
Conversely, let $x$ be such that $d_S(x)$ is $1$. 
By~$i)$ the  integer $x$ must be in $S$. 
As $x=0+x$ is always a decomposition of $x$ as a sum of two elements in $S$, we obtain $D_S(x)=\{0\}$.
If there exist $y$ and $z$ in $S$ such that $y\leq z$ and $x=y+z$ hold then $y$ lies in~$D_S(x)$. 
This implies $y=0$ and $z=x$. 
Hence $x$ is irreducible in~$S$.
\end{proof}

We note  that $0$ is never irreducible despite the fact $d_S(0)$ is $1$ for all numerical semigroups $S$.

We now explain how to compute the $S$-decomposition numbers function of a numerical semigroup  from that of its father.

@@

\begin{proof}
Let $y$ be in $\NN\setminus\{0\}$. We have 
\[
D_{S^x}(y)=\{z \in S^x\ |\ y-z\in S^x\ \text{and}\  2z\leq y\}.
\]
We construct two sets as follow :
\begin{align*}
E&=\{z\in S^x\ |\ y-z=x\ \text{and}\ 2z\leq y\}\\
F&=\begin{cases}
\{x\}&\text{if $y-x\in S$ and $2x\leq y$},\\
\emptyset&\text{otherwise}.
\end{cases}
\end{align*}
We have the relation $D_{S}(y)=D_{S^x}(y)\sqcup E\sqcup F$ and so $d_S(y)$ is equal to $d_{S^x}(y)+\card(E)+\card(F)$
and $D_{S^x}(y)$ is a subset of $D_S(y)$. We start by determining $E$. 
The relation $y-z=x$ implies $z=y-x$ and $2z\leq y$ becomes $y\leq 2x$. Thus we obtain
\[
E=\begin{cases}
\{y-x\}&\text{if $y-x\in S^x$ and $y\leq 2x$}\\
\emptyset&\text{otherwise}.
\end{cases}
\] 
The relation $y-x\in S^x$ together with $y\leq2x$ is equivalent to $y-x\in S$ together with $y<2x$.
Therefore we obtain
\[
E=\begin{cases}
\{y-x\}&\text{if $y-x\in S$ and $y< 2x$}\\
\emptyset&\text{otherwise}.
\end{cases}
\] 
Assume $y-x$ lies in $S$.  If $y<2x$ holds, we have $E=\{y-x\}$ and $F=\emptyset$. Otherwise, we obtain $E=\emptyset$ and $F=\{x\}$. 
Then if $y-x\in S$ holds, we have $\card(E)+\card(F)=1$ and so $d_{S}(y)$ is equal to $d_{S^x}(y)+1$.
If $y-x$ does not belong to $S$, then $E$ and $F$ are empty implying $d_{S}(y)=d_{S^x}(y)$. As the condition $y-x\in S$ is equivalent to $y\geq x$ together with $d_S(y-x)>0$, by Proposition~\ref{P:Dandd} the result holds.
\end{proof}

@@

In Algorithm~\ref{A:ExpRec} we do not specify how to compute $c(S)$, $g(S)$ and $m(S)$  from $S$ neither how to test if an integer is irreducible. 
It also misses  the characterization of $S^x$ from $S$.
These items depend heavily of the representation of $S$.
Our choice is to use the $S$-decomposition numbers function. 
The first task is to use a finite set of such numbers to characterize the whole semigroup.

\begin{prop}
\label{P:delta}
Let $G$ be an integer and $S$ be a numerical semigroup of genus $g\leq G$.
Then $S$ is entirely described by the vector $\delta_S=(d_S(0),...,d_S(3\,G)) \in \NN^{3G+1}$. More precisely we can obtain $c(S), g(S), m(S)$ and $\Irr(S)$ from $\delta_S$.
\end{prop}

\begin{proof}
By Proposition~\ref{P:Res} $iii)$ we have  the relation $c(S)\leq 2g(S)$ and so the $S$-decomposition number of~$c(S)$ occurs in $\delta_S$.
Since $c(S)$ is equal to $1+\max(\NN\setminus S)$ for $S\not=\NN$, Proposition~\ref{P:delta} implies 
\[
c(S)=1+\max\{i\in\{0,...,3G\},\ d_S(i)=0\}.
\]
As all elements of $\NN\setminus S$ are smaller than $c(S)$, their $S$-decomposition numbers are in $\delta_S$ and we obtain
\[
g(S)=\card\{i\in\{0,...,3G\},\ d_S(i)=0\}.
\]
By Proposition~\ref{P:Res} $ii)$, the relation $m(S)\leq g(S)+1$ holds. 
This implies that the $S$-decomposition number of $m(S)$ appears in $\delta_S$ : 
\[
m(S)=\min\{i\in\{0,...,3G\},\ d_S(i)>0\}.
\]
By Proposition~\ref{P:Res},  all irreducible elements are smaller than $c(S)+m(S)-1$, which is itself smaller than~$3G$.
Hence, Proposition~\ref{P:Dandd} gives
\[
\Irr(S)=\{i \in\{0,...,3G\},\ d_S(i)=1\}.
\]
\end{proof}

Even though it is quite simple, the computation of $c(S),m(S)$ and $g(S)$ from $\delta_S$ has a non negligible cost.
We represent a numerical semigroup $S$ of genus $g\leq G$ by $(c(S),g(S),c(S),\delta_S)$. 
In an algorithmic context, if the variable \texttt{S} stands for a numerical semigroups  we use:

@@

Even though there are asymptotically faster algorithms than the one presented here,
thank to careful optimizations, we were able to compute $n_g$ for much larger
genuses than before. This is due to the fact that our algorithm is particularly
well suited for the current processor architecture. In particular, it allows
to use parallelism at various scales (parallel branch exploration,
vectorization)...

To get the greatest speed from modern processors, we used several optimization
tricks, which we will elaborate in the following section:

@@

The source code of our algorithm is available here \cite{code}.
\subsection{Vectorization}

Assume for example that we want to construct the tree~$\mathcal{T}_{100}$
of all numerical semigroups of genus smaller than $100$.  In this case, the
representation of numerical semigroups given in Section~\ref{S:DecNumber} uses
decomposition numbers of integers smaller than~$300$.  By Lemma~\ref{L:RepN},
such a decomposition number is smaller than $151$ and requires~$1$ byte of
memory.  Thus at each \textbf{for} step of Algorithm~\textsc{Son}, the CPU
actually works on $1$ byte. However current CPUs usually work on $8$ bytes and
even on $16$~bytes using vector extensions. 
The first optimization uses this point.

To go further we must specify that the array of decomposition numbers in the
representation of a semigroup corresponds to consecutive bytes in memory. In
the \textbf{for} loop of Algorithm~\textsc{Son} we may imagine two cursors:
the first one, denoted \texttt{src} pointing to the memory byte of
$\texttt{S.d[0]}$ and the second one, denoted \texttt{dst} pointing to the
memory byte $\texttt{T.d[y]}$.  Using these two cursors, Lines \tt{10} to
\tt{14} of Algorithm~\ref{A:Son} can be rewritten as follows:
